---
title: "Plotting F2 heatmap"
author: "Mary T. Yohannes"
date: "11/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pheatmap)
library(grid)
library(reshape2)
library(dplyr)
```

# In order to run the code below, you first need to download the pairwaise doubleton count table, produced in HGDP+1kGP tutorial nb4, locally. Make sure to update the input and output file paths in the code below accordingly.  

## This markdown shows how to make a sample-level F2 heatmap and calculate F2 related mean & SD statistics 

# 1) Set up matrix for heatmap
```{r}
# read-in table with doubleton counts 
ds <- read.csv(file = '~/Desktop/Broad/alicia/HGDP_TGP/fst_and_f2/doubleton_count.csv', sep = "\t") # 5707131 rows × 3 columns

# exclude rows where a sample is set against itself 
ds_not <- ds[!(ds$sample1 == ds$sample2), ]

# sanity check 
dim(ds)[1] - 3378 == dim(ds_not)[1] # TRUE

# reverse columns and rename col names 
reverse_ds <- ds_not[c("sample2", "sample1", "count")] 
names(reverse_ds) <- names(ds_not) 

# join original and reversed tables together 
joined_ds <- rbind(ds_not,reverse_ds)

# set up final table 
final_ds <- rbind(ds[(ds$sample1 == ds$sample2), ], joined_ds)

# sanity check 
dim(final_ds)[1] == dim(joined_ds)[1] + 3378 # TRUE

# cast final table (3378 by 3379 - the extra one column is the sample IDs)
cor_mat <- dcast(final_ds, sample1~sample2, value.var='count')

# update row names with sample IDs  
rownames(cor_mat) <- cor_mat[,1]

# remove the sample ID column and convert data frame into a correlation matrix (3380 by 3380)
cor_mat <- as.matrix(cor_mat[-1])
```


# 2) Set up annotation data frame and colors for heatmap - region for row and population for column
```{r}
# read-in metadata from gnomAD and select only the desired columns  
hgdp_tgp_unfil <- read.delim('~/Desktop/Broad/alicia/HGDP_TGP/gnomad_meta_for_plots.tsv', header=T, sep='\t') 

# set up df for annotation  
hgdp_tgp <- hgdp_tgp_unfil[hgdp_tgp_unfil$s %in% unique(ds$sample1),] # subset metadata to only samples with doubletons 
hgdp_tgp <- hgdp_tgp[c("s", "population", "genetic.region")] # select only sample ID, population and region columns 
rownames(hgdp_tgp) <- hgdp_tgp[,1] # set sample IDs as rownames 
hgdp_tgp <- hgdp_tgp[-1] # remove the sample ID column 
colnames(hgdp_tgp) <- c("Population", "Region") # rename columns 
hgdp_tgp_reg <- within(hgdp_tgp, rm(Population)) # only keep the region column  
hgdp_tgp_pop <- within(hgdp_tgp, rm(Region)) # only keep the population column 

### set color for heat map annotation - obtain from metadata itself for consistency across all plots
# region 
reg_colr_df <- unique(hgdp_tgp_unfil[c("genetic.region", "continent.colors")])
reg_colr <- reg_colr_df %>% pull(continent.colors, genetic.region)

# population within each region  
pop_colr_df <- unique(hgdp_tgp_unfil[c("population", "pop.colors")])
pop_colr <- pop_colr_df %>% pull(pop.colors, population)

# convert into a list 
annot_colr <- list(Region = reg_colr, Population = pop_colr) 
```


# 3a) Trial run on a subsetted ds - since the whole heatmap is dense, printing it crashes the markdown 
```{r}
trial_mat <- cor_mat[1:500,1:500] # subset dataset for trial purposes 
trial_log <- log10(trial_mat+1) # take the log10 since the count values have a big range; +1 to avoid values that are ~0
make_space <-vector(mode="character", length = length(rownames(trial_log))) # used to make space between plot and legend 

# heatmap 
setHook("grid.newpage", function() pushViewport(viewport(x=1,y=1,width=.9, height=.9, name="vp", just=c("right","top"))), action="prepend")
legn_range <- head(seq(0, max(trial_log), .5), -1) # legend range 
pheatmap(trial_log, annotation_row = hgdp_tgp_pop, annotation_col = hgdp_tgp_reg, angle_col = 90, annotation_legend = F, annotation_colors = annot_colr, border_color=NA, treeheight_row=0,height=7, width=9, show_rownames = T, show_colnames = F, legend_breaks = c(legn_range, max(trial_log)), legend_labels = round(c((10^legn_range)-1, (10^(max(trial_log)))-1), 0), labels_row = make_space)
setHook("grid.newpage", NULL, "replace")
grid.text("Samples", x = 0.45, y=0.08, gp=gpar(fontsize=13)) # xlabel
grid.text("Samples", x=-0.02, rot=90, gp=gpar(fontsize=13)) # ylabel
```

# 3b) Plot heatmap with original dataset - takes ~7min to run; saves as a png when you run the code chunk 
```{r}
mat_log <- log10(cor_mat+1)
make_space <- vector(mode="character", length = length(rownames(mat_log))) # used to make space between plot and legend 
legn_range <- head(seq(0, max(mat_log), .5), -1) # legend range 
png(file="f2_heatmap.png", width = 9, height = 9, units = 'in', res = 300) # uncomment this line and comment the next one (pdf) to save heatmap as png 
#pdf(file="f2_heatmap.pdf", width = 9, height = 9) # save as pdf 
setHook("grid.newpage", function() pushViewport(viewport(x=1,y=1,width=.9, height=.9, name="vp", just=c("right","top"))), action="prepend")
pheatmap(mat_log, annotation_row = hgdp_tgp_pop, annotation_col = hgdp_tgp_reg, annotation_colors = annot_colr, annotation_legend = F, angle_col = 90, border_color=NA, treeheight_row=0, height=7, width=9, show_rownames = T, show_colnames = F, legend_breaks = c(legn_range, max(mat_log)), legend_labels = round(c((10^legn_range)-1, (10^(max(mat_log)))-1), 0), labels_row = make_space)
setHook("grid.newpage", NULL, "replace")
grid.text("Samples", x = 0.5, y=0.08, gp=gpar(fontsize=13)) # xlabel
grid.text("Samples", x=-0.02, rot=90, gp=gpar(fontsize=13)) # ylabel
dev.off() # close file ; x = 0.45, y=0.14
```

# 3c) Plot only the upper triangle of the f2 heatmap - saves as a png when you run the code chunk
```{r}
# order the rows and columns of the correlation matrix according to the clustering from the heatmap and set the lower triangle to NA values (keeping the original dendogram)
mat_log <- log10(cor_mat+1)
rnames <- rownames(mat_log)
hc <-  hclust(dist(mat_log))
mat_log2 <- mat_log[hc$order, hc$order]
mat_log2[lower.tri(mat_log2)] <- NA
mat_log2 <- mat_log2[rnames, rnames]

make_space <-vector(mode="character", length = length(rownames(mat_log2))) # used to make space between plot and legend 
legn_range <- head(seq(0, max(mat_log2, na.rm = T), .5), -1) # legend range 
png(file="f2_heatmap_uppertri.png", width = 9, height = 9, units = 'in', res = 300) # save heatmap as png 
setHook("grid.newpage", function() pushViewport(viewport(x=1,y=1,width=.9, height=.9, name="vp", just=c("right","top"))), action="prepend")
pheatmap(mat_log2, cluster_col = hc, cluster_row = hc, annotation_row = hgdp_tgp_pop, annotation_col = hgdp_tgp_reg, angle_col = 90, annotation_legend = F, annotation_colors = annot_colr, border_color=NA, treeheight_row=0,height=7, width=9, show_rownames = T, show_colnames = F, legend_breaks = c(legn_range, max(mat_log)), legend_labels = round(c((10^legn_range)-1, (10^(max(mat_log)))-1), 0), labels_row = make_space, na_col="white")
setHook("grid.newpage", NULL, "replace")
grid.text("Samples", x = 0.45, y=0.08, gp=gpar(fontsize=13)) # xlabel
grid.text("Samples", x=-0.02, rot=90, gp=gpar(fontsize=13)) # ylabel
dev.off() # close file 
```


# 4a) Get stats for manuscript - set up data
```{r}
# read-in sample ID, population, genetic region, and project columns from the metadata 
hgdp_tgp_meta <- read.delim('~/Desktop/Broad/alicia/HGDP_TGP/gnomad_meta_for_plots.tsv', header=T, sep='\t') 

# merge the dataset with the annotation dataframe 
# first join for pop1, then pop2, then rename the added columns
ds_with_meta <-  left_join(ds, hgdp_tgp_meta, by=c("sample1" = "s")) %>% left_join(hgdp_tgp_meta, by=c("sample2" = "s")) %>% rename(sample1_pop = population.x, sample1_GR = genetic.region.x, sample1_project = project.x, sample2_pop = population.y, sample2_GR = genetic.region.y, sample2_project = project.y)
```

# 4b) Calculate mean and standard deviation statistics 
```{r}
# create new columns based on certain conditions 
ds_with_meta <- ds_with_meta %>% mutate(
  # check if samples are from the same population  
  same_pop = case_when(sample1_pop == sample2_pop ~ "Yes",
                       sample1_pop != sample2_pop ~ "No" ),
  # check if samples are from the same genetic region 
  same_GR = case_when(sample1_GR == sample2_GR ~ "Yes",
                      sample1_GR != sample2_GR ~ "No"))

#### mean average of doubleton #####

## pairs of individuals within a population 
ds_with_meta %>% filter(same_pop == "Yes") %>% summarise(mean(count)) # mean = 51.62388	; 51.82778 (old)	

## pairs of individuals within AFR populations
ds_with_meta %>% filter(same_pop == "Yes", sample1_GR == "AFR") %>% summarise(mean(count)) # mean = 76.3766

## pairs of individuals within out-of-Africa populations
ds_with_meta %>% filter(same_pop == "Yes", sample1_GR != "AFR") %>% summarise(mean(count)) # mean = 43.43508; 43.73861 (old)	

## pairs of individuals across populations within a genetic region
ds_with_meta %>% filter(same_GR == "Yes", same_pop == "No") %>% summarise(mean(count), sd(count)) # mean = 6.837925; 6.789911 (old), sd = 19.09835; 18.30799 (old)	

## pairs of individuals across populations from different genetic regions 
ds_with_meta %>% filter(same_GR == "No", same_pop == "No") %>% summarise(mean(count), sd(count)) # mean = 0.7969933, sd = 1.778934		
```


# 5) Look more into the sample pairs that have high # of doubletons 
```{r}
# look through 
ds_with_meta[order(ds_with_meta$count, decreasing = TRUE),] %>% select(sample1, sample2, count, sample1_pop, sample2_pop, same_GR) %>% filter(same_GR == "No") %>% slice_head(n = 15) 

# the highest doubleton counts are seen among the San samples: LP6005443-DNA_G08; LP6005441-DNA_B11; HGDP01029; LP6005441-DNA_A11; HGDP00992; SS6004473; Top 15 pairs are among San samples - doubleton counts between 14717 - 24374. The highest doubleton count across populations is 4068 which is between the BantuSouthAfrica and the San which are both in the same genetic region. Top doubleton counts are among the same popualtion groups. 
# Need to do more deep dive on the amount of doubleton counts among samples that are from different genetic regions.

# The top 15 highest doubleton counts are among individual pairs from the San population with doubleton counts between 14,717 - 24,374. Very few doubletons are shared among pairs of individuals across populations within a geographical/genetic region (mu=6.846.79, sd=19.118.31) with the highest doubleton count being 4,068 between individuals from BantuSouthAfrica and San which are both AFR populations., Evenand even fewer doubletons are shared among pairs of individuals across populations from different genetic regions (mu=0.8, sd=1.78) with the highest doubleton count being 638 between a pair of individuals from CDX and BEB which are EAS and CSA populations, respectively. 

```




---------------
###### JUNK CODE #######

### 5) Figure out which populations are interspersed in heatmap 
```{r}
mat_log <- log10(cor_mat+1)

# generate heatmap and assign to a variable 
p_heatmap <- pheatmap(mat_log, annotation_row = hgdp_tgp_pop, annotation_col = hgdp_tgp_reg, annotation_colors = annot_colr, annotation_legend = F, angle_col = 90, border_color=NA, treeheight_row=0, height=7, width=9, show_rownames = T, show_colnames = F)

# plot heatmap dendrogram 
plot(p_heatmap$tree_row) 

# re-order original data (samples) to match ordering in heatmap
samples_id <- rownames(cor_mat[p_heatmap$tree_row[["order"]],]) 

# convert vector in to a dataframe
samples_df <- data.frame(matrix(unlist(samples_id), nrow=length(samples_id), byrow=TRUE)) 

# add column name
colnames(samples_df) <- "Samples" 

# obtain region and population info for each sample using the metadata df used for heatmap annotation 
hgdp_tgp2 <- hgdp_tgp
hgdp_tgp2$Samples <- rownames(hgdp_tgp2) # convert rownames in to a column

# merge the two dfs while keeping the order 
# by going through the region column in the merged df, we can figure out which populations are interspersed 
df_merge <- inner_join(samples_df,hgdp_tgp2)
df_merge

write.csv(df_merge, "~/Desktop/Broad/alicia/HGDP_TGP/plots_all/f2_interspersed_samples.csv", row.names=FALSE)

# height of each branch
dend <- as.dendrogram(p_heatmap$tree_row)
plot(dend)

f2_node_height <- get_nodes_attr(dend, "height")
height_df <- data.frame(matrix(unlist(f2_node_height), nrow=length(f2_node_height), byrow=TRUE)) # convert to df

#colnames(mat_df) <- "Samples"
write.csv(height_df, "~/Desktop/Broad/alicia/HGDP_TGP/plots_all/f2_dendrogram_heights.csv", row.names=FALSE)
```




