---
title: "Plotting F2 heatmap"
author: "Mary T. Yohannes"
date: "11/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pheatmap)
library(grid)
library(reshape2)
library(dplyr)
```

## This markdown shows how to make a sample-level F2 heatmap and calculate F2 related mean & SD statistics 

### 1) Set up matrix for heatmap
```{r}
# read-in table with doubleton counts 
ds <- read.csv(file = '~/Desktop/Broad/alicia/HGDP_TGP/fst_and_f2/doubleton_count.csv', sep = "\t") # 5707131 rows × 3 columns

# exclude rows where a sample is set against itself 
ds_not <- ds[!(ds$sample1 == ds$sample2), ]

# sanity check 
dim(ds)[1] - 3378 == dim(ds_not)[1] # TRUE

# reverse columns and rename col names 
reverse_ds <- ds_not[c("sample2", "sample1", "count")] 
names(reverse_ds) <- names(ds_not) 

# join original and reversed tables together 
joined_ds <- rbind(ds_not,reverse_ds)

# set up final table 
final_ds <- rbind(ds[(ds$sample1 == ds$sample2), ], joined_ds)

# sanity check 
dim(final_ds)[1] == dim(joined_ds)[1] + 3378 # TRUE

# cast final table (3378 by 3379 - the extra one column is the sample IDs)
cor_mat <- dcast(final_ds, sample1~sample2, value.var='count')

# update row names with sample IDs  
rownames(cor_mat) <- cor_mat[,1]

# remove the sample ID column and convert data frame into a correlation matrix (3380 by 3380)
cor_mat <- as.matrix(cor_mat[-1])

# make a heatmap - takes a bit time to run 
#pheatmap(cor_mat)
```

### 2) Set up annotation data frame and colors for heat map - region for row and population for column
```{r}
# read-in metadata from gnomAD and select only the desired columns - this file is either the original metadata from the drive (downloaded the excel sheet as a tsv) or one sent by Alicia to me and Zan 
hgdp_tgp_unfil <- read.delim('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/gnomad_meta_v1.tsv', header=T, sep='\t') %>% select(s, project_meta.title, starts_with('hgdp_tgp_meta')) 

# some sample IDs have 'v3.1::' prefix as done in gnomAD so stripping that in order to match correctly with the pc score files 
hgdp_tgp_unfil$s <- gsub('v3.1::', '', hgdp_tgp_unfil$s)

# set up df for annotation  
hgdp_tgp <- hgdp_tgp_unfil[hgdp_tgp_unfil$s %in% unique(ds$sample1),] # subset metadata to only samples with doubletons 
hgdp_tgp <- hgdp_tgp[c("s", "hgdp_tgp_meta.Population", "hgdp_tgp_meta.Genetic.region")] # select only sample ID, population and region columns 
rownames(hgdp_tgp) <- hgdp_tgp[,1] # set sample IDs as rownames 
hgdp_tgp <- hgdp_tgp[-1] # remove the sample ID column 
colnames(hgdp_tgp) <- c("Population", "Region") # rename columns 
hgdp_tgp_reg <- within(hgdp_tgp, rm(Population)) # only keep the region column  
hgdp_tgp_pop <- within(hgdp_tgp, rm(Region)) # only keep the population column 

### set color for heat map annotation - obtain from metadata itself for consistency across all plots
# region 
reg_colr_df <- unique(hgdp_tgp_unfil[c("hgdp_tgp_meta.Genetic.region", "hgdp_tgp_meta.Continent.colors")])
reg_colr <- reg_colr_df %>% pull(hgdp_tgp_meta.Continent.colors, hgdp_tgp_meta.Genetic.region)
#annot_reg_colr <- list(Region = reg_colr) # convert into a list  

# population within each region  
pop_colr_df <- unique(hgdp_tgp_unfil[c("hgdp_tgp_meta.Population", "hgdp_tgp_meta.Pop.colors")])
pop_colr <- pop_colr_df %>% pull(hgdp_tgp_meta.Pop.colors, hgdp_tgp_meta.Population)
annot_colr <- list(Region = reg_colr, Population = pop_colr) # convert into a list 
```

### 3a) Trial run on a subsetted ds
```{r}
trial_mat <- cor_mat[1:500,1:500] # subset dataset for trial purposes 
trial_log <- log10(trial_mat+1) # take the log10 since the count values have a big range; +1 to avoid values that are ~0
make_space <-vector(mode="character", length = length(rownames(trial_log))) # used to make space between plot and legend 

# close to final heatmap 
setHook("grid.newpage", function() pushViewport(viewport(x=1,y=1,width=.9, height=.9, name="vp", just=c("right","top"))), action="prepend")
legn_range <- head(seq(0, max(trial_log), .5), -1) # legend range 
pheatmap(trial_log, annotation_row = hgdp_tgp_pop, annotation_col = hgdp_tgp_reg, angle_col = 90, annotation_legend = F, annotation_colors = annot_colr, border_color=NA, treeheight_row=0,height=7, width=9, show_rownames = T, show_colnames = F, legend_breaks = c(legn_range, max(trial_log)), legend_labels = round(c((10^legn_range)-1, (10^(max(trial_log)))-1), 0), labels_row = make_space)
setHook("grid.newpage", NULL, "replace")
grid.text("Samples", x = 0.45, y=0.08, gp=gpar(fontsize=13)) # xlabel
grid.text("Samples", x=-0.02, rot=90, gp=gpar(fontsize=13)) # ylabel
```

### 3b) Plot heatmap with original dataset - takes a bit time to run (~7min); save as pdf or png
```{r}
mat_log <- log10(cor_mat+1)
make_space <-vector(mode="character", length = length(rownames(mat_log))) # used to make space between plot and legend 
legn_range <- head(seq(0, max(mat_log), .5), -1) # legend range 
#png(file="f2_heatmap.png", width = 9, height = 9, units = 'in', res = 300) # uncomment this line and comment the next one (pdf) to save heatmap as png 
pdf(file="f2_heatmap.pdf", width = 9, height = 9) # save as pdf 
setHook("grid.newpage", function() pushViewport(viewport(x=1,y=1,width=.9, height=.9, name="vp", just=c("right","top"))), action="prepend")
pheatmap(mat_log, annotation_row = hgdp_tgp_pop, annotation_col = hgdp_tgp_reg, annotation_colors = annot_colr, annotation_legend = F, angle_col = 90, border_color=NA, treeheight_row=0, height=7, width=9, show_rownames = T, show_colnames = F, legend_breaks = c(legn_range, max(mat_log)), legend_labels = round(c((10^legn_range)-1, (10^(max(mat_log)))-1), 0), labels_row = make_space)
setHook("grid.newpage", NULL, "replace")
grid.text("Samples", x = 0.5, y=0.08, gp=gpar(fontsize=13)) # xlabel
grid.text("Samples", x=-0.02, rot=90, gp=gpar(fontsize=13)) # ylabel
dev.off() # close file ; x = 0.45, y=0.14
```

# 4a) Get stats for manuscript - set up data
```{r}
# read-in sample ID, population, genetic region, and project columns from the metadata 
hgdp_tgp_meta <- read.delim('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/gnomad_meta_v1.tsv', header=T, sep='\t') %>% select(s, hgdp_tgp_meta.Population, hgdp_tgp_meta.Genetic.region, project_meta.title) 

# some sample IDs have 'v3.1::' prefix as done in gnomAD so stripping that in order to match correctly with the pc score files 
hgdp_tgp_meta$s <- gsub('v3.1::', '', hgdp_tgp_meta$s)

# merge the dataset with the annotation dataframe 
# first join for pop1, then pop2, then rename the added columns
ds_with_meta <-  left_join(ds, hgdp_tgp_meta, by=c("sample1" = "s")) %>% left_join(hgdp_tgp_meta, by=c("sample2" = "s")) %>% rename(sample1_pop = hgdp_tgp_meta.Population.x, sample1_GR = hgdp_tgp_meta.Genetic.region.x, sample1_project = project_meta.title.x, sample2_pop = hgdp_tgp_meta.Population.y, sample2_GR = hgdp_tgp_meta.Genetic.region.y, sample2_project = project_meta.title.y)
```

# 4b) Calculate mean and standard deviation statistics 
```{r}
# create new columns based on certain conditions 
ds_with_meta <- ds_with_meta %>% mutate(
  # check if samples are from the same population  
  same_pop = case_when(sample1_pop == sample2_pop ~ "Yes",
                       sample1_pop != sample2_pop ~ "No" ),
  # check if samples are from the same genetic region 
  same_GR = case_when(sample1_GR == sample2_GR ~ "Yes",
                      sample1_GR != sample2_GR ~ "No"))

#### mean average of doubleton #####

## pairs of individuals within a population 
ds_with_meta %>% filter(same_pop == "Yes") %>% summarise(mean(count)) # mean = 51.82778	

## pairs of individuals within AFR populations
ds_with_meta %>% filter(same_pop == "Yes", sample1_GR == "AFR") %>% summarise(mean(count)) # mean = 76.3766	

## pairs of individuals within out-of-Africa populations
ds_with_meta %>% filter(same_pop == "Yes", sample1_GR != "AFR") %>% summarise(mean(count)) # mean = 43.73861	

## pairs of individuals across populations within a genetic region
ds_with_meta %>% filter(same_GR == "Yes", same_pop == "No") %>% summarise(mean(count), sd(count)) # mean = 6.789911, sd = 18.30799	

## pairs of individuals across populations from different genetic regions 
ds_with_meta %>% filter(same_GR == "No", same_pop == "No") %>% summarise(mean(count), sd(count)) # mean = 0.7969933, sd = 1.778934		
```