---
title: "Obtain Failed Samples & Plot Ratio"
author: "Mary T. Yohannes"
date: "1/26/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("dplyr")
library("ggplot2")
```


#### set working directory to where the tsv file is located and import it
```{r}
setwd("~/Desktop/Broad/alicia/HGDP_TGP")

ds_tmp <- read.table("HGDP+TGP_gnomAD_subset_metadata.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE, fill = TRUE)

# set the sample column as row names
ds <- ds_tmp[,-1]
rownames(ds) <- ds_tmp[,1]

#dim(ds) # 4150 by 183
#summary(ds)
```


### Task 1: obtain samples (within the 5 popns) that failed the QC filters
#### extraction of populations and columns of interest 
```{r}
# subset data to only the populations and columns we are interested in - Note: Melanesian is aka Bougainville
# all populations had at least one "TRUE" value for the filter columns so only needed the sample IDs, population names, and the filter columns
filtered_out <- c("BiakaPygmy", "MbutiPygmy", "Melanesian", "Papuan", "San") # populations that were filtered out by gnomAD

all_popn <- data.frame() # empty data frame to store the subset data 
for (i in filtered_out){
  one_popn <- ds[ds$hgdp_tgp_meta.Population == i, ][,c(175, 129:156)] # obtain the columns of interest for each of the 5 popns
  all_popn <- rbind.data.frame(all_popn, one_popn) # make one dataset 
}
```

#### obtain which and how many filters each sample failed (had true value for the QC filters)
```{r}
# list of columns that had "TRUE" value/s - failed the gnomAD filters 
all_popn$failed_filters <- apply(all_popn==TRUE, 1, FUN = function(x) toString (names(x)[x]))
# remove NA from the list of column names 
all_popn$failed_filters <- gsub("\\NA, ", "", all_popn$failed_filters)

# total number of "TRUE" values per sample - total number of failed samples    
all_popn$failed_filters_count <- rowSums(all_popn == TRUE, na.rm = T)

# sanity check
#length(unlist(strsplit(all_popn$failed_filters[1], split=","))) == all_popn$failed_filters_count[1]

# rearrange the columns - move the last two columns towards the beginning of the data frame 
all_popn <- all_popn[c(1, 31, 30, 2:29)]
```

#### save the result 
```{r}
# only write out the columns we are interested in
subset_all_popn <- all_popn[c(1:3)]

# save as csv file 
write.csv(subset_all_popn, file = "samples_that_failed.csv")
```


### Task 2: obtain number of samples within popns that failed the QC filters 
#### number of samples in each popn 
```{r}
t_freq <- ds %>% count(hgdp_tgp_meta.Population) 

# edit the column names 
colnames(t_freq) <- c("population", "num_of_samples")
```

#### number of samples per popn that failed each filter  
```{r}
# first, group the dataset by population and then within each population, add up the number of samples that had a value of "TRUE" for each filter column (failed that filter)
t_filter <- ds %>% group_by(hgdp_tgp_meta.Population)  %>% summarise_at(129:156, ~sum(.==TRUE, na.rm = TRUE))

#~sum(.==TRUE, na.rm = TRUE) is the same as funs(sum(.==TRUE, na.rm = TRUE))

#ds %>% group_by(hgdp_tgp_meta.Population)  %>% summarise(p = sum(sample_filters.fail_n_transversion_residual == TRUE, na.rm = TRUE)) # does what the above code does but only for one filter column - for each population, add up the number of samples that failed the "sample_filters.fail_n_transversion_residual" filter

# sanity check 
#table(ds %>% group_by(hgdp_tgp_meta.Population)  %>% summarise(p = sum(sample_filters.sex_aneuploidy == TRUE, na.rm = TRUE)) %>% select(p) == ds %>% group_by(hgdp_tgp_meta.Population)  %>% summarise_at(129:156, ~sum(.==TRUE, na.rm = TRUE)) %>% select(sample_filters.sex_aneuploidy))
```

#### additional data manipulation 
```{r}
# merge the two dataframes 
t_final <- merge(t_freq, t_filter, by.x = "population", by.y = "hgdp_tgp_meta.Population")

# add a column to indicate the 5 populations that were filtered out by the gnomAD QC
t_final$failed_gnomAD <- t_final$population %in% filtered_out
```

#### save final table as csv file 
```{r}
write.csv(t_final, file = "failed_filters_population_level.csv", row.names = FALSE)
```


### Task 3: plot the ratios and save as one pdf file
#### set up data for plotting  
```{r warning = FALSE}
# for each filter column, calculate the ratio between the number of samples that failed and the total number of samples in the population. Then, select only the population, failed_gnomAD and the ratio columns for plotting  
t_ratio <- t_final %>% mutate_at(vars(sample_filters.sex_aneuploidy:sample_filters.fail_n_transversion_residual), funs(P = ./t_final$num_of_samples)) %>% select(c(population, ends_with("_P"), failed_gnomAD))

# list to hold the plots in 
plot_list <- list()
# generate scatter plots of ratios for each filter column across all populations colored by gnomAD failure 
for (i in colnames(t_ratio)[2:29]){
    filter_col <- gsub(".*[.]([^.]+)[_].*", "\\1", i) # set up column names correctly 
    p <- ggplot(t_ratio, aes_string(x=i, y="population", colour="failed_gnomAD")) + 
      geom_point() +  
      labs(x=filter_col, title=paste("Population vs", filter_col, sep=" ")) + 
      theme(axis.text.y = element_text(size = 7)) 
    plot_list[[i]] <- p
}
```

#### save plots as one pdf 
```{r results = 'hide'}
# create a pdf file where each page is a separate plot
pdf("gnomAD_filters_scatter_plots.pdf")
for (i in 1:28) {
    print(plot_list[[i]])
}
dev.off()
```


