---
title: "Plotting PCA Plots - with and without outliers"
author: "Mary T. Yohannes"
date: "11/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# import needed libraries 
library(tidyverse)
library(RColorBrewer)
library(plotly)
library(cowplot)
library(grid)
library(gridExtra)
```

## This markdown is divided into two sections: PCA plots used to detect outlier samples (with outliers) and plots generated after the outliers were removed (without outliers). Global and subcontinental plots with PCs 1 through 8 were generated and written out for both sections. We first set up the plotting functions: 

# 1) Functions 
### 1a) function to generate a single PCA plot 
```{r}
# function to plot a single [simple] PCA plot
single_pca_plot <- function(region = c('GLOBAL', 'AFR', 'EUR', 'AMR', 'EAS', 'CSA', 'OCE', 'MID'),
                            pc_scores, 
                            which_pc1, 
                            which_pc2, 
                            bottom=F){
  
  pc_scores <- na.omit(pc_scores) # remove possible NA values - # CHMI_CHMI3_WGS2 has no info in metadata so treated as an outlier
  
  # plot specifics 
  pca_plot <- ggplot(aes=(text='s')) +
    geom_point(size=2) +
    theme_classic() +
    labs(shape="Projects") +
    theme(text = element_text(size=14, color='black'),
          legend.title = element_text(size = 12),
          legend.text = element_text(size = 9),
          legend.key.size = unit(0.7, "lines")) +
    guides(shape = guide_legend(override.aes = list(size = 0.85)))
  
  if(bottom) {
    pca_plot <- pca_plot +
      theme(axis.text = element_text(color='black'),
            legend.title = element_text(size = 12),
            legend.text = element_text(size = 9),
            legend.key.size = unit(0.7, "lines"),
            legend.position="bottom") +
      guides(fill=guide_legend(ncol=1,byrow=FALSE, title.position="top"),
             shape=guide_legend(ncol=1,byrow=FALSE, title.position="top"))
  }
  
  # GLOBAL PCA
  if (region == 'GLOBAL'){
    
    # obtain region colors from the metadata for consistency across all plots
    region_color <- unique(pc_scores[c("hgdp_tgp_meta.Genetic.region", "hgdp_tgp_meta.Continent.colors")]) %>% pull(hgdp_tgp_meta.Continent.colors, hgdp_tgp_meta.Genetic.region)

    # add information to plot
    pca_plot <- pca_plot +
      geom_point(pc_scores, mapping = aes_string(x=which_pc1, y=which_pc2, color='hgdp_tgp_meta.Genetic.region', shape='project_meta.title')) + 
      scale_color_manual(values=region_color, labels=unique(pc_scores$hgdp_tgp_meta.Genetic.region), name='Regions') 
    
    # SUBCONTINENTAL PCA
    } else if (region %in% c('AFR', 'EUR', 'AMR', 'EAS', 'CSA', 'OCE', 'MID')) {
      
      # assign colors to populations within each region  
      pop_count <- length(unique(pc_scores$hgdp_tgp_meta.Population))
      pop_color <- colorRampPalette(brewer.pal(7, "Set1"))(pop_count)
      
      # add information to plot
      pca_plot <- pca_plot +
        geom_point(pc_scores, mapping = aes_string(x=which_pc1, y=which_pc2, color='hgdp_tgp_meta.Population', shape='project_meta.title')) + 
        scale_color_manual(values=pop_color, labels=unique(pc_scores$hgdp_tgp_meta.Population), name='Populations') 
        
      # in case user doesn't specify 'region' argument when running the function  
      } else{
        print("Wrong entry for the 'region' argument. Please choose from the following options: 'GLOBAL', 'AFR', 'EUR', 'AMR', 'EAS', 'CSA', 'OCE', or 'MID'")
        }
  return(pca_plot)
  }
```


### 1b) function to generate plots PC1 through PC8, organize them in a grid and save as a pdf  
```{r}
pca_plots_in_grid <- function(region, outlier_status){
  
  # make sure the string entries are in the right format
  region <- toupper(region)
  outlier_status <- tolower(outlier_status)
  
  # read-in pc scores to plot 
  region_pc_scores <- read.table(gzfile(paste0('pca/', outlier_status, '/scores/', region, '_scores_', outlier_status, '.txt.bgz')), header=T)
  
  # read-in gnomAD metadata and select desired columns 
  hgdp_tgp_metadata <- read.delim('~/Desktop/Broad/alicia/pca_subcont/gnomad_meta_v1.tsv', header=T, sep='\t') %>% select(s, project_meta.title, starts_with('hgdp_tgp_meta')) 
  
  # some sample IDs have 'v3.1::' prefix as done in gnomAD so stripping that in order to match correctly with the pc score files 
  hgdp_tgp_metadata$s <- gsub('v3.1::', '', hgdp_tgp_metadata$s)
  
  # join scores and metadata to add additional sample info
  pc_scores_meta <- region_pc_scores %>% left_join(hgdp_tgp_metadata) %>% arrange(hgdp_tgp_meta.Population)
  
  # plot 
  pca_1_2 <- single_pca_plot(region, pc_scores_meta, 'PC1', 'PC2')
  pca_1_2_bottom <- single_pca_plot(region, pc_scores_meta, 'PC1', 'PC2', TRUE) 
  pca_3_4 <- single_pca_plot(region, pc_scores_meta, 'PC3', 'PC4')
  pca_5_6 <- single_pca_plot(region, pc_scores_meta, 'PC5', 'PC6')
  pca_7_8 <- single_pca_plot(region, pc_scores_meta, 'PC7', 'PC8')
  all_pca <- plot_grid(pca_1_2 + theme(legend.position="none"), pca_3_4 + theme(legend.position="none"), pca_5_6 + theme(legend.position="none"), pca_7_8 + theme(legend.position="none"), nrow=1) # put all 4 plots together 
  legend <- get_legend(pca_1_2)
  all_pca_add_legend <- plot_grid(all_pca, legend, rel_widths = c(4.8, 1)) # add legend
  all_pca_add_title <- arrangeGrob(all_pca_add_legend, top=textGrob(region, x = 0, hjust = 0)) # add title 
  
  # save plots in grid as pdf 
  ggsave(paste0('pca/', outlier_status, '/plots/', region, '_pca_1to8_', outlier_status, '.pdf'), all_pca_add_title, width=15, height=4)
  
  return(list(all_pca_add_title, pca_1_2_bottom))
}
```


# 2) Plot PCA
### 2a) With Outliers  
```{r}
# GLOBAL PCA 
GLOBAL_pca <- pca_plots_in_grid('global', 'with_outliers')

# Subcontinental PCA
regions <- c('AFR', 'EUR', 'AMR', 'EAS', 'CSA', 'OCE', 'MID')
for (i in 1:length(regions)){
  assign(paste0(regions[i], '_pca'), pca_plots_in_grid(regions[i], 'with_outliers'))
}

# Combine all plots together 
ALL_pca <- plot_grid(GLOBAL_pca[[1]], AFR_pca[[1]], AMR_pca[[1]], CSA_pca[[1]], EAS_pca[[1]], EUR_pca[[1]], MID_pca[[1]], OCE_pca[[1]], ncol=1, align='v')

# Save as pdf and png
ggsave('pca/with_outliers/plots/ALL_pca_1to8_with_outliers.pdf', ALL_pca, width=15, height=28)
ggsave('pca/with_outliers/plots/ALL_pca_1to8_with_outliers.png', ALL_pca, width=15, height=28)
```

### 2b) Without Outliers  
```{r}
# GLOBAL PCA 
GLOBAL_pca <- pca_plots_in_grid('global', 'without_outliers')

# Subcontinental PCA
regions <- c('AFR', 'EUR', 'AMR', 'EAS', 'CSA', 'OCE', 'MID')
for (i in 1:length(regions)){
  assign(paste0(regions[i], '_pca'), pca_plots_in_grid(regions[i], 'without_outliers'))
}

# Combine all plots together 
ALL_pca <- plot_grid(GLOBAL_pca[[1]], AFR_pca[[1]], AMR_pca[[1]], CSA_pca[[1]], EAS_pca[[1]], EUR_pca[[1]], MID_pca[[1]], OCE_pca[[1]], ncol=1, align='v')

# Save as pdf and png
ggsave('pca/without_outliers/plots/ALL_pca_1to8_without_outliers.pdf', ALL_pca, width=15, height=28)
ggsave('pca/without_outliers/plots/ALL_pca_1to8_without_outliers.png', ALL_pca, width=15, height=28)
```

