---
title: "Plotting Subcontinental PCA before and after outliers were removed (UPDATED)"
author: "Mary T. Yohannes"
date: "3/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RColorBrewer)
library(plotly)
library(cowplot)
```

# 1) set working directory and import metadata
```{r}
setwd("~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp")
hgdp_tgp <- read.delim('~/Desktop/Broad/alicia/pca_subcont/gnomad_meta_v1.tsv', header=T, sep='\t') %>%
  dplyr::select(s, project_meta.title, starts_with('hgdp_tgp_meta')) 

# some sample IDs have 'v3.1::' prefix as done in gnomAD so stripping that in order to match correctly with the pc score files 
hgdp_tgp$s <- gsub('v3.1::', '', hgdp_tgp$s)
```

# Before Removing Outlier Samples
# 2a) function for plotting only pc1 & 2 and save as a pdf 
```{r}
load_pop_make_pca_plot <- function(pop_name) {
  # import pc scores 
  ref_unrel <- read.table(gzfile(paste0('UPDATED_scores_for_pca_preoutlier/subcont_pca_', pop_name, '_scores.txt.bgz')), header=T)
  proj_rel <- read.table(gzfile(paste0('UPDATED_scores_for_pca_preoutlier/subcont_pca_', pop_name, '_projected_scores.txt.bgz')), header=T)
  
  # combine the ref and proj data into one dataset and then join that to the metadata to add additional info for those samples 
  pca_subcont <- bind_rows(ref_unrel, proj_rel) %>%
    left_join(hgdp_tgp)
  
  nb.cols <- length(unique(pca_subcont$hgdp_tgp_meta.Population)) # number of subcontinental regions within each region/continent 
  mycolors <- colorRampPalette(brewer.pal(7, "Set1"))(nb.cols) # set color palettes 
  
  # plot pca 
  p_subcont <- ggplot(pca_subcont, aes(x=PC1, y=PC2, color=hgdp_tgp_meta.Population, shape=project_meta.title, text=s)) +
    geom_point() +
    theme_classic() +
    labs(shape='Project') +
    scale_color_manual(values=mycolors, name='Population') +
    labs(title=paste(pop_name, "UPDATED PCA (before removing outliers)"))
  
  ggsave(paste0('UPDATED_pca_plots_preoutlier/UPDATED_preoutlier_subcont_pca_1n2_', pop_name, '.pdf'), p_subcont) # save plot to pdf 
  ggplotly(p_subcont, tooltip='text') # make plot interactive 
}
```

# 2b) plot subcontinental pcas (only pc1 & 2) for each region 
```{r}
# AFR PCA 
load_pop_make_pca_plot('AFR') 

# AMR PCA 
load_pop_make_pca_plot('AMR') 

# CSA PCA 
load_pop_make_pca_plot('CSA') 

# EAS PCA 
load_pop_make_pca_plot('EAS') 

# EUR PCA 
load_pop_make_pca_plot('EUR') 

# MID PCA 
load_pop_make_pca_plot('MID') 
	
# OCE PCA 
load_pop_make_pca_plot('OCE') 
```

# 3a) function to plot subcont PCAs in a grid upto the 8th PC 
```{r}
single_pca_plot <- function(pca_subcont, first_PC, second_PC, pop_name, bottom=F) {
  nb.cols <- length(unique(pca_subcont$hgdp_tgp_meta.Population))
  mycolors <- colorRampPalette(brewer.pal(7, "Set1"))(nb.cols)
  is_tgp <- unique(pca_subcont$hgdp_tgp_meta.Population) %in% subset(pca_subcont, hgdp_tgp_meta.Project=='1000 Genomes')$hgdp_tgp_meta.Population
  myshapes <- ifelse(is_tgp, 16, 4)
  
  p_subcont <- ggplot(pca_subcont, aes_string(x=first_PC, y=second_PC, color='hgdp_tgp_meta.Population', shape='hgdp_tgp_meta.Population', text='s')) +
    geom_point(size=2) +
    theme_classic() +
    scale_shape_manual(values=myshapes, labels=unique(pca_subcont$hgdp_tgp_meta.Population), name='Population') +
    scale_color_manual(values=mycolors, labels=unique(pca_subcont$hgdp_tgp_meta.Population), name='Population') +
    labs(title=pop_name) +
    theme(text = element_text(size=14, color='black'),
          legend.title = element_text(size = 12), 
          legend.text = element_text(size = 9),
          legend.key.size = unit(0.7, "lines")) +
    guides(shape = guide_legend(override.aes = list(size = 0.85)))
  if(bottom) {
    p_subcont <- p_subcont +
      theme(axis.text = element_text(color='black'),
            legend.title = element_text(size = 12), 
            legend.text = element_text(size = 9),
            legend.key.size = unit(0.7, "lines"),
            legend.position="bottom") +
      guides(fill=guide_legend(ncol=1,byrow=FALSE, title.position="top"),
             shape=guide_legend(ncol=1,byrow=FALSE, title.position="top"))
  }
  return(p_subcont)
}


load_pop_make_pca_plot <- function(pop_name) {
  # subcont PCA 
  ref_unrel <- read.table(gzfile(paste0('UPDATED_scores_for_pca_preoutlier/subcont_pca_', pop_name, '_scores.txt.bgz')), header=T)
  proj_rel <- read.table(gzfile(paste0('UPDATED_scores_for_pca_preoutlier/subcont_pca_', pop_name, '_projected_scores.txt.bgz')), header=T)
  
  # combine the ref and proj data into one dataset and then join that to the metadata to add additional info for those samples 
  pca_subcont <- bind_rows(ref_unrel, proj_rel) %>%
    left_join(hgdp_tgp) %>%
    arrange(hgdp_tgp_meta.Population)
  
  p_1_2 <- single_pca_plot(pca_subcont, 'PC1', 'PC2', pop_name)
  p_1_2_bottom <- single_pca_plot(pca_subcont, 'PC1', 'PC2', pop_name, TRUE) 
  p_3_4 <- single_pca_plot(pca_subcont, 'PC3', 'PC4', pop_name)
  p_5_6 <- single_pca_plot(pca_subcont, 'PC5', 'PC6', pop_name)
  p_7_8 <- single_pca_plot(pca_subcont, 'PC7', 'PC8', pop_name)
  p_subcont <- plot_grid(p_1_2 + theme(legend.position="none"), p_3_4 + theme(legend.position="none"), 
                         p_5_6 + theme(legend.position="none"), p_7_8 + theme(legend.position="none"), nrow=1)
  legend <- get_legend(p_1_2)
  p_subcont_legend <- plot_grid(p_subcont, legend, rel_widths = c(4, 1))
  
  ggsave(paste0('UPDATED_pca_plots_preoutlier/UPDATED_preoutlier_subcont_pca_1to8', pop_name, '.pdf'), p_subcont_legend, width=15, height=4)
  
  ggplotly(p_subcont_legend, tooltip='text')
  return(list(p_subcont_legend, p_1_2_bottom))
}
```

# 3b) plot subcontinental pcas in a grid upto the 8th PC 
```{r}
p_AFR <- load_pop_make_pca_plot('AFR') # outlier: NA20314 (PC1/2), NA20299 (PC3/4), HG01880 (PC6), HG01881 (PC6)
p_AMR <- load_pop_make_pca_plot('AMR')
p_CSA <- load_pop_make_pca_plot('CSA') # outlier: HGDP00130 (PC1/2), HGDP00013 (PC4&5), HGDP00150 (PC5), HGDP00029
p_EAS <- load_pop_make_pca_plot('EAS') # outlier: HGDP01298 (PC3), HGDP01303 (PC4&5), LP6005443-DNA_B02 (PC5), HGDP01300 (PC6)
p_EUR <- load_pop_make_pca_plot('EUR') # outlier: HG01628 (PC5), HG01629 (PC5), HG01630 (PC5), HG01694 (PC6), HG01696 (PC6)
p_MID <- load_pop_make_pca_plot('MID') # outlier: HGDP00621, HGDP01270, HGDP01271
p_OCE <- load_pop_make_pca_plot('OCE') # outlier: HGDP00554 - not actually an outlier when looking at global PCs

# all pc plots together in a grid - 4 per region (upto the 8th pc)
p_region_outliers <- plot_grid(p_AFR[[1]], p_AMR[[1]], p_CSA[[1]], p_EAS[[1]], p_EUR[[1]], p_MID[[1]], p_OCE[[1]], 
                               labels=LETTERS[1:7], ncol=1, align='v')

# save as pdf and png
ggsave('UPDATED_pca_plots_preoutlier/UPDATED_preoutlier_regions_pca_1to8.pdf', p_region_outliers, height=28, width=15)
ggsave('UPDATED_pca_plots_preoutlier/UPDATED_preoutlier_regions_pca_1to8.png', p_region_outliers, height=15, width=12)

# just pc1 & 2 in a grid for all the regions - if the plots that are individually produced in 2a and 2b were put together in a grid  
p_region_outliers_1_2 <- plot_grid(p_AFR[[2]], p_AMR[[2]], p_CSA[[2]], p_EAS[[2]], p_EUR[[2]], p_MID[[2]], p_OCE[[2]], nrow=1, align='h', labels=LETTERS[5:11])

# save as pdf and png
#ggsave('UPDATED_pca_plots_preoutlier/UPDATED_preoutlier_regions_pca_1n2.pdf', p_region_outliers_1_2, height=28, width=15)
#ggsave('UPDATED_pca_plots_preoutlier/UPDATED_preoutlier_regions_pca_1n2.png', p_region_outliers_1_2, height=15, width=12)
```


# 4a) global PCA - only pc1 & 2 
```{r}
ref <- read.table(gzfile('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/UPDATED_scores_for_pca_preoutlier/global_scores.txt.bgz'), header=T)
proj <- read.table(gzfile('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/UPDATED_scores_for_pca_preoutlier/global_projected_scores.txt.bgz'), header=T)

# combine the ref and proj data into one dataset and then join that to the metadata to add additional info for those samples
pca <- bind_rows(ref, proj) %>%
  left_join(hgdp_tgp)

# plot pca 
p <- ggplot(pca, aes(x=PC1, y=PC2, color=hgdp_tgp_meta.Genetic.region, shape=project_meta.title, text=s)) +
  geom_point() +
  theme_classic() +
  labs(title="UPDATED Global PCA (before removing outliers)")

ggsave('UPDATED_pca_plots_preoutlier/UPDATED_global_pca.pdf', p) # save plot as pdf 
ggplotly(p, tooltip='text') # make plot interactive 
```

# 4b) function to plot global pca in a grid upto the 8th PC 
```{r}
sin_pca_plot <- function(pca_subcont, first_PC, second_PC, pop_name, bottom=F) {
  nb.cols <- length(unique(pca_subcont$hgdp_tgp_meta.Population))
  mycolors <- colorRampPalette(brewer.pal(7, "Set1"))(nb.cols)
  is_tgp <- unique(pca_subcont$hgdp_tgp_meta.Population) %in% subset(pca_subcont, hgdp_tgp_meta.Project=='1000 Genomes')$hgdp_tgp_meta.Population
  myshapes <- ifelse(is_tgp, 16, 4)
  
  p_subcont <- ggplot(pca_subcont, aes_string(x=first_PC, y=second_PC, color='hgdp_tgp_meta.Population', shape='hgdp_tgp_meta.Population', text='s')) +
    geom_point(size=2) +
    theme_classic() +
    scale_shape_manual(values=myshapes, labels=unique(pca_subcont$hgdp_tgp_meta.Population), name='Population') +
    scale_color_manual(values=mycolors, labels=unique(pca_subcont$hgdp_tgp_meta.Population), name='Population') +
    labs(title='global') +
    theme(text = element_text(size=14, color='black'),
          legend.title = element_text(size = 12), 
          legend.text = element_text(size = 9),
          legend.key.size = unit(0.7, "lines")) +
    guides(shape = guide_legend(override.aes = list(size = 0.85)))
  if(bottom) {
    p_subcont <- p_subcont +
      theme(axis.text = element_text(color='black'),
            legend.title = element_text(size = 12), 
            legend.text = element_text(size = 9),
            legend.key.size = unit(0.7, "lines"),
            legend.position="bottom") +
      guides(fill=guide_legend(ncol=1,byrow=FALSE, title.position="top"),
             shape=guide_legend(ncol=1,byrow=FALSE, title.position="top"))
  }
  return(p_subcont)
}
```

# 4c) plot global pcs and save as pdf 
````{r}
ref_unrel <- read.table(gzfile('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/UPDATED_scores_for_pca_preoutlier/global_scores.txt.bgz'), header=T)
proj_rel <- read.table(gzfile('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/UPDATED_scores_for_pca_preoutlier/global_projected_scores.txt.bgz'), header=T)

# combine the ref and proj data into one dataset and then join that to the metadata to add additional info for those samples 
pca_subcont <- bind_rows(ref_unrel, proj_rel) %>%
  left_join(hgdp_tgp) %>%
  arrange(hgdp_tgp_meta.Population)
  
p_1_2 <- sin_pca_plot(pca_subcont, 'PC1', 'PC2')
p_1_2_bottom <- sin_pca_plot(pca_subcont, 'PC1', 'PC2', TRUE) 
p_3_4 <- sin_pca_plot(pca_subcont, 'PC3', 'PC4')
p_5_6 <- sin_pca_plot(pca_subcont, 'PC5', 'PC6')
p_7_8 <- sin_pca_plot(pca_subcont, 'PC7', 'PC8')
p_subcont <- plot_grid(p_1_2 + theme(legend.position="none"), p_3_4 + theme(legend.position="none"), 
                         p_5_6 + theme(legend.position="none"), p_7_8 + theme(legend.position="none"), nrow=1)
legend <- get_legend(p_1_2)
p_subcont_legend <- plot_grid(p_subcont, legend, rel_widths = c(4, 1))
  
ggsave(paste0('UPDATED_pca_plots_preoutlier/UPDATED_global_pca_1to8.pdf'), p_subcont_legend, width=15, height=4)
  
ggplotly(p_subcont_legend, tooltip='text')
```

# ----------------------------------------------------------------------------------------

# After Removing Outlier Samples
# 5a) function for plotting only pc1 & 2 and save as a pdf 
```{r}
load_pop_make_pca_plot <- function(pop_name) {
  # import pc scores 
  ref_unrel <- read.table(gzfile(paste0('UPDATED_scores_for_pca_postoutlier/subcont_pca_', pop_name, '_scores.txt.bgz')), header=T)
  proj_rel <- read.table(gzfile(paste0('UPDATED_scores_for_pca_postoutlier/subcont_pca_', pop_name, '_projected_scores.txt.bgz')), header=T)
  
  # combine the ref and proj data into one dataset and then join that to the metadata to add additional info for those samples 
  pca_subcont <- bind_rows(ref_unrel, proj_rel) %>%
    left_join(hgdp_tgp)
  
  nb.cols <- length(unique(pca_subcont$hgdp_tgp_meta.Population)) # number of subcontinental regions within each region/continent 
  mycolors <- colorRampPalette(brewer.pal(7, "Set1"))(nb.cols) # set color palettes 
  
  # plot pca 
  p_subcont <- ggplot(pca_subcont, aes(x=PC1, y=PC2, color=hgdp_tgp_meta.Population, shape=project_meta.title, text=s)) +
    geom_point() +
    theme_classic() +
    labs(shape='Project') +
    scale_color_manual(values=mycolors, name='Population') +
    labs(title=paste(pop_name, "UPDATED PCA (after removing outliers)"))
  
  ggsave(paste0('UPDATED_pca_plots_postoutlier/UPDATED_postoutlier_subcont_pca_1n2_', pop_name, '.pdf'), p_subcont) # save plot to pdf 
  ggplotly(p_subcont, tooltip='text') # make plot interactive 
}
```

# 5b) plot subcontinental pcas (only pc1 & 2) for each region 
```{r}
# AFR PCA 
load_pop_make_pca_plot('AFR') 

# AMR PCA 
load_pop_make_pca_plot('AMR') 

# CSA PCA 
load_pop_make_pca_plot('CSA') 

# EAS PCA 
load_pop_make_pca_plot('EAS') 

# EUR PCA 
load_pop_make_pca_plot('EUR') 

# MID PCA 
load_pop_make_pca_plot('MID') 
	
# OCE PCA 
load_pop_make_pca_plot('OCE') 
```

# 6a) function to plot subcont PCAs in a grid upto the 8th PC 
```{r}
single_pca_plot <- function(pca_subcont, first_PC, second_PC, pop_name, bottom=F) {
  nb.cols <- length(unique(pca_subcont$hgdp_tgp_meta.Population))
  mycolors <- colorRampPalette(brewer.pal(7, "Set1"))(nb.cols)
  is_tgp <- unique(pca_subcont$hgdp_tgp_meta.Population) %in% subset(pca_subcont, hgdp_tgp_meta.Project=='1000 Genomes')$hgdp_tgp_meta.Population
  myshapes <- ifelse(is_tgp, 16, 4)
  
  p_subcont <- ggplot(pca_subcont, aes_string(x=first_PC, y=second_PC, color='hgdp_tgp_meta.Population', shape='hgdp_tgp_meta.Population', text='s')) +
    geom_point(size=2) +
    theme_classic() +
    scale_shape_manual(values=myshapes, labels=unique(pca_subcont$hgdp_tgp_meta.Population), name='Population') +
    scale_color_manual(values=mycolors, labels=unique(pca_subcont$hgdp_tgp_meta.Population), name='Population') +
    labs(title=pop_name) +
    theme(text = element_text(size=14, color='black'),
          legend.title = element_text(size = 12), 
          legend.text = element_text(size = 9),
          legend.key.size = unit(0.7, "lines")) +
    guides(shape = guide_legend(override.aes = list(size = 0.85)))
  if(bottom) {
    p_subcont <- p_subcont +
      theme(axis.text = element_text(color='black'),
            legend.title = element_text(size = 12), 
            legend.text = element_text(size = 9),
            legend.key.size = unit(0.7, "lines"),
            legend.position="bottom") +
      guides(fill=guide_legend(ncol=1,byrow=FALSE, title.position="top"),
             shape=guide_legend(ncol=1,byrow=FALSE, title.position="top"))
  }
  return(p_subcont)
}


load_pop_make_pca_plot <- function(pop_name) {
  # subcont PCA 
  ref_unrel <- read.table(gzfile(paste0('UPDATED_scores_for_pca_postoutlier/subcont_pca_', pop_name, '_scores.txt.bgz')), header=T)
  proj_rel <- read.table(gzfile(paste0('UPDATED_scores_for_pca_postoutlier/subcont_pca_', pop_name, '_projected_scores.txt.bgz')), header=T)
  
  # combine the ref and proj data into one dataset and then join that to the metadata to add additional info for those samples 
  pca_subcont <- bind_rows(ref_unrel, proj_rel) %>%
    left_join(hgdp_tgp) %>%
    arrange(hgdp_tgp_meta.Population)
  
  p_1_2 <- single_pca_plot(pca_subcont, 'PC1', 'PC2', pop_name)
  p_1_2_bottom <- single_pca_plot(pca_subcont, 'PC1', 'PC2', pop_name, TRUE) 
  p_3_4 <- single_pca_plot(pca_subcont, 'PC3', 'PC4', pop_name)
  p_5_6 <- single_pca_plot(pca_subcont, 'PC5', 'PC6', pop_name)
  p_7_8 <- single_pca_plot(pca_subcont, 'PC7', 'PC8', pop_name)
  p_subcont <- plot_grid(p_1_2 + theme(legend.position="none"), p_3_4 + theme(legend.position="none"), 
                         p_5_6 + theme(legend.position="none"), p_7_8 + theme(legend.position="none"), nrow=1)
  legend <- get_legend(p_1_2)
  p_subcont_legend <- plot_grid(p_subcont, legend, rel_widths = c(4, 1))
  
  ggsave(paste0('UPDATED_pca_plots_postoutlier/UPDATED_postoutlier_subcont_pca_1to8', pop_name, '.pdf'), p_subcont_legend, width=15, height=4)
  
  ggplotly(p_subcont_legend, tooltip='text')
  return(list(p_subcont_legend, p_1_2_bottom))
}
```

# 6b) plot subcontinental pcas in a grid upto the 8th PC 
```{r}
p_AFR <- load_pop_make_pca_plot('AFR') # outlier: NA20314 (PC1/2), NA20299 (PC3/4), HG01880 (PC6), HG01881 (PC6)
p_AMR <- load_pop_make_pca_plot('AMR')
p_CSA <- load_pop_make_pca_plot('CSA') # outlier: HGDP00130 (PC1/2), HGDP00013 (PC4&5), HGDP00150 (PC5), HGDP00029
p_EAS <- load_pop_make_pca_plot('EAS') # outlier: HGDP01298 (PC3), HGDP01303 (PC4&5), LP6005443-DNA_B02 (PC5), HGDP01300 (PC6)
p_EUR <- load_pop_make_pca_plot('EUR') # outlier: HG01628 (PC5), HG01629 (PC5), HG01630 (PC5), HG01694 (PC6), HG01696 (PC6)
p_MID <- load_pop_make_pca_plot('MID') # outlier: HGDP00621, HGDP01270, HGDP01271
p_OCE <- load_pop_make_pca_plot('OCE') # outlier: HGDP00554 - not actually an outlier when looking at global PCs

# all pc plots together in a grid - 4 per region (upto the 8th pc)
p_region_outliers <- plot_grid(p_AFR[[1]], p_AMR[[1]], p_CSA[[1]], p_EAS[[1]], p_EUR[[1]], p_MID[[1]], p_OCE[[1]], 
                               labels=LETTERS[1:7], ncol=1, align='v')

# save as pdf and png
ggsave('UPDATED_pca_plots_postoutlier/UPDATED_postoutlier_regions_pca_1to8.pdf', p_region_outliers, height=28, width=15)
ggsave('UPDATED_pca_plots_postoutlier/UPDATED_postoutlier_regions_pca_1to8.png', p_region_outliers, height=15, width=12)

# just pc1 & 2 in a grid for all the regions - if the plots that are individually produced in 2a and 2b were put together in a grid  
p_region_outliers_1_2 <- plot_grid(p_AFR[[2]], p_AMR[[2]], p_CSA[[2]], p_EAS[[2]], p_EUR[[2]], p_MID[[2]], p_OCE[[2]], nrow=1, align='h', labels=LETTERS[5:11])

# save as pdf and png
#ggsave('UPDATED_pca_plots_postoutlier/UPDATED_postoutlier_regions_pca_1n2.pdf', p_region_outliers_1_2, height=28, width=15)
#ggsave('UPDATED_pca_plots_postoutlier/UPDATED_postoutlier_regions_pca_1n2.png', p_region_outliers_1_2, height=15, width=12)
```

# 7a) global PCA - only pc1 & 2 
```{r}
ref <- read.table(gzfile('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/UPDATED_scores_for_pca_postoutlier/global_scores.txt.bgz'), header=T)
proj <- read.table(gzfile('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/UPDATED_scores_for_pca_postoutlier/global_projected_scores.txt.bgz'), header=T)

# combine the ref and proj data into one dataset and then join that to the metadata to add additional info for those samples
pca <- bind_rows(ref, proj) %>%
  left_join(hgdp_tgp)

# plot pca 
p <- ggplot(pca, aes(x=PC1, y=PC2, color=hgdp_tgp_meta.Genetic.region, shape=project_meta.title, text=s)) +
  geom_point() +
  theme_classic() +
  labs(title="UPDATED Global PCA (after removing outliers)")

ggsave('UPDATED_pca_plots_postoutlier/UPDATED_global_pca.pdf', p) # save plot as pdf 
ggplotly(p, tooltip='text') # make plot interactive 
```

# 7b) function to plot global pca in a grid upto the 8th PC 
```{r}
sin_pca_plot <- function(pca_subcont, first_PC, second_PC, pop_name, bottom=F) {
  nb.cols <- length(unique(pca_subcont$hgdp_tgp_meta.Population))
  mycolors <- colorRampPalette(brewer.pal(7, "Set1"))(nb.cols)
  is_tgp <- unique(pca_subcont$hgdp_tgp_meta.Population) %in% subset(pca_subcont, hgdp_tgp_meta.Project=='1000 Genomes')$hgdp_tgp_meta.Population
  myshapes <- ifelse(is_tgp, 16, 4)
  
  p_subcont <- ggplot(pca_subcont, aes_string(x=first_PC, y=second_PC, color='hgdp_tgp_meta.Population', shape='hgdp_tgp_meta.Population', text='s')) +
    geom_point(size=2) +
    theme_classic() +
    scale_shape_manual(values=myshapes, labels=unique(pca_subcont$hgdp_tgp_meta.Population), name='Population') +
    scale_color_manual(values=mycolors, labels=unique(pca_subcont$hgdp_tgp_meta.Population), name='Population') +
    labs(title='global') +
    theme(text = element_text(size=14, color='black'),
          legend.title = element_text(size = 12), 
          legend.text = element_text(size = 9),
          legend.key.size = unit(0.7, "lines")) +
    guides(shape = guide_legend(override.aes = list(size = 0.85)))
  if(bottom) {
    p_subcont <- p_subcont +
      theme(axis.text = element_text(color='black'),
            legend.title = element_text(size = 12), 
            legend.text = element_text(size = 9),
            legend.key.size = unit(0.7, "lines"),
            legend.position="bottom") +
      guides(fill=guide_legend(ncol=1,byrow=FALSE, title.position="top"),
             shape=guide_legend(ncol=1,byrow=FALSE, title.position="top"))
  }
  return(p_subcont)
}
```

# 7c) plot global pcs and save as pdf 
````{r}
ref_unrel <- read.table(gzfile('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/UPDATED_scores_for_pca_postoutlier/global_scores.txt.bgz'), header=T)
proj_rel <- read.table(gzfile('~/Desktop/Broad/alicia/HGDP_TGP/hgdp_tgp/UPDATED_scores_for_pca_postoutlier/global_projected_scores.txt.bgz'), header=T)

# combine the ref and proj data into one dataset and then join that to the metadata to add additional info for those samples 
pca_subcont <- bind_rows(ref_unrel, proj_rel) %>%
  left_join(hgdp_tgp) %>%
  arrange(hgdp_tgp_meta.Population)
  
p_1_2 <- sin_pca_plot(pca_subcont, 'PC1', 'PC2')
p_1_2_bottom <- sin_pca_plot(pca_subcont, 'PC1', 'PC2', TRUE) 
p_3_4 <- sin_pca_plot(pca_subcont, 'PC3', 'PC4')
p_5_6 <- sin_pca_plot(pca_subcont, 'PC5', 'PC6')
p_7_8 <- sin_pca_plot(pca_subcont, 'PC7', 'PC8')
p_subcont <- plot_grid(p_1_2 + theme(legend.position="none"), p_3_4 + theme(legend.position="none"), 
                         p_5_6 + theme(legend.position="none"), p_7_8 + theme(legend.position="none"), nrow=1)
legend <- get_legend(p_1_2)
p_subcont_legend <- plot_grid(p_subcont, legend, rel_widths = c(4, 1))
  
ggsave(paste0('UPDATED_pca_plots_postoutlier/UPDATED_global_pca_1to8.pdf'), p_subcont_legend, width=15, height=4)
  
ggplotly(p_subcont_legend, tooltip='text')
```
